preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-08 Manuscript/manuscript.Rmd
Polish results

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table S1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table S3).

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason being two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We extended our method and developed *preciseTAD*, an R package which leverages an optimally built RF to precisely predict the genomic coordinates of TAD and loop boundaries at base resolution.

We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1).

In our ML framework boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2). The number of non-boundary regions highly outnumbered boundary regions, with results consistent for loop boundaries as well. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

@@

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8052                       |                             15468                           |                535363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7676                       |                             14253                           |                267682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4670                       |                             8363                            |                107073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        Peakachu       |       10   kb     |                        16185                       |                          21421                              |                267682               |        0.14            |
Table: TAD/loop boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on the GM12878 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types when predicting TAD boundary regions.

@@

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome that was left out of training and reserved for testing. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, under our machine learning framework we were able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained (using the optimal characteristics outlined above) on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7846 TAD and 15838 chromatin loop boundaries on GM12878, as well as 8883 TAD and 11016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 2). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix (Figure 1A). <!--Likewise, Arrowhead often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two domains are inherently conjoined by only one predicted boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] -->Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages 4 known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites. This methodology likely leads to many more false positives, resulting in inflated number of chromatin loops.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more enriched for CTCF sites than either Arrowhead or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more enriched for other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were found to be statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5D, Supplementary Figure 7).

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5E).

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results shown on Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.

@@

**Supplementary Table 2.** TAD/loop boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on the k562 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

**Supplementary Figure 1.** Density curves illustrate the normalization of distances before (red) and after (blue) performing a $log_{2}$ transformation across differing bin sizes including 10 kb, 25 kb, 50 kb, and 100 kb for both the (A) GM12878 and (B) K562 cell lines respectively. Distances are measured as the number of bases from the center of a genomic bin to the center of the nearest genomic annotation. Each density curve represents individual genomic annotation.

@@

**Supplementary Figure 9.** Barplots illustrating the normalized overlap between known molecular drivers of 3D chromatin and flanked predicted TAD boundaries identified by preciseTAD (using Arrowhead ground truth boundaries) on GM12878 and for chromosome 22. Normalized overlap was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted (on chr22). Boundaries were flanked by 5 kb resolution.

>>>>

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 3).

@@

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1). We developed *preciseTAD*, an R package which leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell type-specific genome annotation data.

In our ML framework boundary regions were defined as regions containing a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 2). The number of non-boundary regions highly outnumbered boundary regions. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

@@

|          Tool         |     Resolution    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:-----------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb      |                         8,052                       |                             15,468                           |                535,363               |           0.03         |
|     Arrowhead         |       10   kb     |                         7,676                       |                             14,253                           |                267,682               |           0.05         |
|     Arrowhead         |       25   kb     |                         4,670                       |                             8,363                            |                107,073               |           0.08         |
|     Arrowhead         |       50   kb     |                         2,349                       |                             4,224                            |                 53,537               |           0.08         |
|     Arrowhead         |      100   kb     |                         1,031                       |                             1,883                            |                 26,768               |           0.07         |
|        Peakachu       |       10   kb     |                        16,185                       |                          21,421                              |                267,682               |        0.14            |
Table: TAD/loop boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on the GM12878 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types when predicting TAD boundary regions.

@@

We sought to further optimize our boundary region prediction models by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Using our optimally built random forest model trained on Arrowhead/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space is used to predict the probability of each base being a boundary. Our method, referred to as *preciseTAD*, uses density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location at base resolution (see Methods).

When built using Arrowhead and Peakachu ground truth boundaries at 5 kb and 10 kb respectively, *preciseTAD* was able to predict a total of 7,846 TAD and 15,838 chromatin loop boundaries on GM12878, as well as 8,883 TAD and 11,016 chromatin loop boundaries on K562. Noticeably, we reported less predicted TAD boundaries at 5 kb than Arrowhead on both cell lines (Table 1, Supplementary Table 2). This can be attributed to Arrowhead's inflation of called TADs, especially at higher resolutions, that often do not correspond to intra-chromosomal contacts featured on a contact matrix. <!--Likewise, Arrowhead often produces nested and overlapped TADs that do not have demarcated boundaries, further exacerbating this inflation. However, under the framework of *preciseTAD*, any two domains are inherently conjoined by only one predicted boundary, leading to comparatively less TAD boundaries. [??? Unclear explanation, so what? preciseTAD is driven by genomic annotations. May it be too stringent thresholds? This all may be for the discussion, if we have a good understanding why] -->Additionally, *preciseTAD* boundaries were visually more likely to demarcate high concentrations of intra-chromosomal contacts as can be seen on a Hi-C contact matrix (Figure 5A). *preciseTAD* also predicted fewer chromatin loop boundaries than Peakachu (Table 1, Supplementary Table 2). This can be contributed to the fact that *preciseTAD* leverages four known drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143, contrasting Peakachu which considers only CTCF sites as true positives. These results suggest that genome annotation-driven predicted boundaries may be more biologically relevant, in contrast to boundaries detected from Hi-C data alone.

Next, we evaluated the biological significance of *preciseTAD* boundaries. Signal heatmaps indicated that our predicted boundaries were more frequently co-localized with CTCF sites than either Arrowhead or Peakachu boundaries alone (Figure 5 B-C). Likewise, *preciseTAD* boundaries were also more co-localized with other molecular drivers of 3D chromatin including RAD21, SMC3, and ZNF143 for both GM12878 and K562 cell lines (Supplementary Figures 5-6). When compared to called TAD or loop boundaries, *preciseTAD* boundaries were statistically significantly closer to the same cell line-specific top ranked TFBS (p<0.001 and p<0.001 versus Arrowhead and Peakachu respectively) (Figure 5D, Supplementary Figure 7). These results indicate that *preciseTAD*-predicted boundaries better reflect the known bioloby of boundary formation.

@@

TAD boundaries are known to be relatively conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. To assess conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu after flanking them by the resolution of the Hi-C data. Boundaries involving Arrowhead were flanked by 5 kb, while boundaries involving Peakachu were flanked by 10 kb. We found that 35% of *preciseTAD* predicted TAD boundaries and 55% of *preciseTAD* predicted loop boundaries were were conserved between cell lines, compared to only 26% and 55% for Arrowhead and Peakachu alone (Figure 5E). The better conservation of *preciseTAD*-predicted boundaries further support the notion of their higher biological relevance.

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model. We found that 95.6% and 95.7% of our predicted TAD boundaries were overlapped by predicted loop boundaries for GM12878 and K562, respectively (Supplementary Figure 8). This is expected as loop boundaries (Peakachu) are more abundant, while TAD boundaries (Arrowhead) represent the narrower level of the 3D chromatin organization. Together with the results on Figure 4, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar protein complexes.

@@

**Supplementary Table 2.** TAD/loop boundary data and class imbalance summaries across resolutions for Arrowhead and Peakachu on the K562 cell line. Class imbalance is calculated as the ratio of total number of unique boundaries to the total number of genomic bins.

@@

**Supplementary Figure 1.** Density curves illustrate the normalization of distances before (red) and after (blue) performing a $log_{2}$ transformation across 10 kb, 25 kb, 50 kb, and 100 kb data resolutions for both the (A) GM12878 and (B) K562 cell lines, respectively. Distances are measured as the number of bases from the center of a genomic bin to the center of the nearest genomic annotation. Each density curve represents individual genomic annotation. [??? All of them? Or, TFBSs only?]

@@

**Supplementary Figure 9.** Barplots illustrating the normalized overlap between known molecular drivers of 3D chromatin and flanked predicted TAD boundaries identified by preciseTAD (using Arrowhead ground truth boundaries) on GM12878 and for chromosome 22. Normalized overlap was calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries that were predicted (on chr22). Boundaries were flanked by 5 kb resolution. [??? Where is the threshold description?]
